RNA-seq from 4 cell populations

### 2021-08-03 deconvolution/fans4/molec_profiling.Rmd
bug fix

# cd /hpc/users/hoffmg01/work/cmc_analysis/src
cd /hpc/users/hoffmg01/work/CMC/covarr-de/deconvolution/fans4
# rm -rf molec_profiling_cache/
ml git pandoc
git pull
# export OMP_NUM_THREADS=1
R

@@

# set color codes for each cell type
colorCodes = c(
  GABA = '#66A61E',
  GLU = '#E6AB02',
  Olig = '#E7298A',
  MgAs = '#7570B3')

@@

contrasts['GABAGLU_vs_Other'] = "(CellTypeGABA + CellTypeGLU)/2 - (CellTypeMgAs + CellTypeOlig)/2"
contrasts['MgAs_vs_Other'] = "CellTypeMgAs - (CellTypeGABA + CellTypeGLU + CellTypeOlig)/3"
contrasts['Olig_vs_Other'] = "CellTypeOlig - (CellTypeGABA + CellTypeGLU + CellTypeMgAs)/3"

@@

# Gene set analysis
```{r zenith.celltype, cache=FALSE}

geneSets_GO = get_GeneOntology("MF")

# Map from Ensembl genes in geneSets_GO to 
# Ensembl names from RNA-seq data 
index = ids2indices( recodeToList(geneSets_GO), rownames(fitBasic))

# filter by size of gene set
n_genes_in = 30
setSize = sapply(index, length)
idxGO = index[(setSize > n_genes_in) & (setSize < 500)]

df_CT_zenith = lapply( colnames(coef(fitBasic)), function(key){
  res = zenith( fitBasic, coef = key, index = idxGO)
  data.table(Geneset = rownames(res), res, coef = key)
})
df_CT_zenith = do.call(rbind, df_CT_zenith)

@@

  df_CT_zenith[(coef == key),][order(p.greater),][1:4,]
  })
df_gs = do.call(rbind, df_gs)
```

  res = zenith( fitBasic, coef = 'CellTypeGLU', index = idxGO)

>>>>

# cd /hpc/users/hoffmg01/work/cmc_analysis/src
cd /hpc/users/hoffmg01/work/CMC/covarr-de/deconvolution/fans4
# rm -rf molec_profiling_cache/
ml git pandoc
git pull
export OMP_NUM_THREADS=2
R

@@

# set color codes for each cell type
colorCodes = c(
  GABA = '#66A61E',
  GLU = '#E6AB02',
  Olig = '#E7298A',
  MgAs = '#7570B3',
  GABAGLU_vs_Other = '#A6A910',
  GABA_vs_GLU = "grey")

@@

contrasts['GABAGLU_vs_Other'] = "(CellTypeGABA + CellTypeGLU)/2 - (CellTypeMgAs + CellTypeOlig)/2"
contrasts['GLU_vs_Other'] = "CellTypeGLU - (CellTypeGABA + CellTypeMgAs + CellTypeOlig)/3"
contrasts['GABA_vs_Other'] = "CellTypeGABA - (CellTypeGLU + CellTypeMgAs + CellTypeOlig)/3"
contrasts['MgAs_vs_Other'] = "CellTypeMgAs - (CellTypeGABA + CellTypeGLU + CellTypeOlig)/3"
contrasts['Olig_vs_Other'] = "CellTypeOlig - (CellTypeGABA + CellTypeGLU + CellTypeMgAs)/3"

@@

# Gene set analysis
```{r zenith.celltype, cache=FALSE, eval=FALSE}

geneSets_GO = get_GeneOntology()

filter_gs = function(fit, gs){
  # Map from Ensembl genes in geneSets_GO to 
  # Ensembl names from RNA-seq data 
  index = ids2indices( recodeToList(gs), rownames(fit))

  # filter by size of gene set
  n_genes_in = 15
  setSize = sapply(index, length)
  idxGO = index[(setSize > n_genes_in) & (setSize < 500)]
  idxGO
}

fit = fitBasic[fitBasic$Amean > -Inf,]
idxGO = filter_gs( fit, geneSets_GO)

df_CT_zenith = lapply( colnames(coef(fitBasic)), function(key){
  res = zenith( fit, coef = key, index = idxGO, inter.gene.cor=0.01)
  data.table(Geneset = rownames(res), res, coef = key)
})
df_CT_zenith = do.call(rbind, df_CT_zenith)

@@

  # results for this coef
  df_sub = df_CT_zenith[(coef == key),][order(p.greater),]

  # remove gene sets with small p-values from other tests
  df_sub2 = df_CT_zenith[(coef != key),][order(p.greater),]
  df_sub2 = df_sub2[,.SD[which.min(p.greater),], by="Geneset"]
  df_sub2 = df_sub2[p.greater > 1e-3,]

  df_sub[!(df_sub$Geneset %in% df_sub2$Geneset),][1:4,]
  })
df_gs = do.call(rbind, df_gs)

df_gs %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE)

df = df_gs[grep("CellType", coef),]
# df = rbind(df, df_gs[grep('GABAGLU_vs_Other|GABA_vs_GLU', coef)])
df[,coef := gsub("CellType", '',coef)]
df$coef = factor(df$coef, names(colorCodes))

ggplot(df, aes(Geneset,coef, -log10(p.greater))) + geom_tile() + coord_flip() + scale_y_continuous(expand=c(0,0)) + theme_classic() + ylab(bquote(-log[10]~p)) + scale_fill_manual(values = c(colorCodes, GABAGLU_vs_Other = "green"), name="Cell type") 
```

```{r compare.celltypes}
zenith_gsa = function (fit, coefs, geneSets, n_genes_min = 10, n_genes_max=1000,...){

    # from GeneSetCollection to list 
    geneSets = recodeToList(geneSets)

    index = ids2indices(geneSets, rownames(fit))
    index = index[sapply(index, length) >= n_genes_min]
    index = index[sapply(index, length) < n_genes_max]

    df_res = lapply(coefs, function(coef) {
        df_res = zenith(fit, coef, index,...)
        data.table(coef = coef, Geneset = rownames(df_res), df_res)
    })
    df_zenith = do.call(rbind, df_res)
    rownames(df_zenith) = c()

    df_zenith$FDR = p.adjust(df_zenith$PValue, "BH")
    df_zenith
}

geneSets_GO = get_GeneOntology("BP")

coefTests = colnames(coef(fitBasic))[c(1, 7:9)]
df_zenith = zenith_gsa(fitBasic[fitBasic$Amean > 0,], coefTests, geneSets_GO, inter.gene.cor=0.01)

df_zenith$FDR = p.adjust(df_zenith$p.greater, "BH")

df = df_zenith[,.SD[order(p.greater)[1:3],],by=c("coef")]
cls = unique(df$Geneset)
df = df_zenith[Geneset %in% df$Geneset,]

nrow = length(unique(df$coef))
ncol = length(unique(df$Geneset))

M = matrix(-log10(df$p.greater), nrow=nrow, ncol=ncol,dimnames=list(unique(df$coef), unique(df$Geneset)), byrow=TRUE)

hcl_col = hclust(dist(M), method="ward.D")
df$coef = factor(df$coef, rownames(M)[hcl_col$order])

hcl_row = hclust(dist(t(M)), method="ward.D")
# df$Geneset = factor(df$Geneset, colnames(M)[hcl_row$order])
df$Geneset = factor(df$Geneset, cls)
df$label = ifelse(df$FDR < 0.05, '*', '')

ggplot(df, aes(coef, Geneset, fill=-log10(p.greater), label=label)) + geom_tile() + theme_classic() + scale_fill_gradient(limits=c(0, NA), low="white", high="red", name=bquote(-log[10]~P)) + theme(aspect.ratio = ncol/nrow, axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) + xlab('') + ylab('') + geom_text(vjust=1, hjust=0.5)
```
